Introduction to the \proglang{R} package \pkg{TDA}

### 2021-03-11 inst/doc/article.Rnw
version 1.7.5

\author{Brittany T. Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria, \and Vincent Rouvreau \\ \small{In collaboration with the CMU TopStat Group}}
\Plainauthor{Brittany Terese Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria, Vincent Rouvreau}

\title{Introduction to the \proglang{R} package \pkg{TDA}}
\Plaintitle{Introduction to the R package TDA}
\Abstract{
We present a short tutorial and introduction to using the \proglang{R} package 
\pkg{TDA}, which provides some tools for Topological Data Analysis.
In particular, it includes implementations of functions that, given some data, provide 
topological information about the underlying space, such as the distance function, the distance to a measure,
the kNN density estimator, the kernel density estimator, and the kernel distance.
The salient topological features of the sublevel sets (or superlevel sets) of these functions
can be quantified with persistent homology. We provide an \proglang{R} interface
for the efficient algorithms of the \proglang{C++} libraries \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}, and \href{ https://bitbucket.org/phat-code/phat }{\pkg{PHAT}}, 
including a function for the persistent homology of the Rips filtration, 
and one for the persistent homology of sublevel
sets (or superlevel sets) of arbitrary functions evaluated over a grid of points.
The significance of the features in the resulting persistence diagrams can be 
analyzed with functions that implement the methods discussed in \cite{fasy2014statistical},
\cite{chazal2014stochastic} and \cite{chazal2014robust}.
The \proglang{R} package \pkg{TDA} also includes the implementation of an algorithm for density 
clustering, which allows us to identify the spatial organization of the probability mass associated to 
a density function and visualize it by means of a dendrogram, the cluster tree. 
}
\Keywords{Topological Data Analysis, Persistent Homology, Density Clustering}

@@

  Jisu Kim\\
  Department of Statistics\\
  Carnegie Mellon University\\
  Website: \url{ http://www.stat.cmu.edu/~jisuk/ } \\
  Email: \email{jisuk1@andrew.cmu.edu}\\

@@

@@

%There is a growing list of software available to compute persistent homology.
%\textsf{\textbf{javaPlex}}\footnote{https://code.google.com/p/javaplex/} is a
%Java software package developed by the Computational Topology workgroup at
%Stanford University;
%\textsf{\textbf{phom}}\footnote{
%http://cran.r-project.org/web/packages/phom/index.html} is an R package written
%by Andrew Tausz;
%\textsf{\textbf{PHAT}}\footnote{https://bitbucket.org/phat-code/phat/} is a C++
%library developed at the Institute of Science and Technology Austria;
%\textsf{\textbf{Dionysus}}\footnote{http://www.mrzv.org/software/dionysus/} is
%a
%C++ library written and maintained by Dmitriy Morozov;
%finally, \textsf{\textbf{GUDHI}}\footnote{https://project.inria.fr/gudhi/} is
%new born project hosted by INRIA and whose goal is the development and
%implementation of new algorithms for geometric understanding in high
%dimensions.
%Preliminary results show that \textsf{\textbf{GUDHI}} outperforms its major
%competitors \textsf{\textbf{PHAT}} and \textsf{\textbf{Dionysus}}; see
%\cite{boissonnat2013compressed}.

This paper is devoted to the presentation of the \proglang{R} package \pkg{TDA},
which provides a user-friendly interface for the efficient algorithms of the 
\proglang{C++} libraries \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}} \citep{gudhi_stpcoh}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}} \citep{dionysus_C}, and \href{ https://bitbucket.org/phat-code/phat }{\pkg{PHAT}} \citep{phat_C}.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Persistent Homology Over a Grid}
%%%%%%%%%%%%%%%%%%%%%%%
In this section, we describe how to use the \code{gridDiag} function to compute the persistent homology of sublevel (and superlevel) sets of the functions described in Section \ref{sec:distances}. 
The function \code{gridDiag} evaluates a given real valued function over a triangulated grid, constructs a filtration of simplices using the values of the function, and computes the persistent homology of the filtration. From version 1.2, \code{gridDiag} works in arbitrary dimension.
The core of the function is written in \proglang{C++} and the user can choose to compute persistence diagrams using either  the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat }{\pkg{PHAT}}.

@@

The \code{ripsDiag} function computes the persistence diagram of the Rips filtration built on top of a point cloud.
The user can choose to compute the Rips filtration using either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}} or \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}. Then for computing the persistence diagram from the Rips filtration, the user can use either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}.
\\The following code generates 60 points from two circles:

@@

The \code{alphaComplexDiag} function computes the Alpha complex filtration built on top of a point cloud, using the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}. Then for computing the persistence diagram from the Alpha complex filtration, the user can use either the C++ library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}.

@@

The \code{alphaShapeDiag} function computes the persistence diagram of the Alpha shape filtration built on top of a point cloud in 3 dimension, using the C++ library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}. Then for computing the persistence diagram from the Alpha shape filtration, the user can use either the C++ library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}. Currently the point data cloud should lie in 3 dimension.

@@

\noindent Once the filtration is computed, the function \code{filtrationDiag} computes the persistence diagram from the filtration. The user can choose to compute the persistence diagram using either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}} or \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}.

@@

Standard metrics for measuring the distance between two persistence diagrams are the bottleneck distance and the $p$th Wasserstein distance \citep{edelsbrunner2010computational}. 
The \pkg{TDA} package includes the functions \code{bottleneck} and \code{wasserstein}, which are R wrappers of the functions ``bottleneck\_distance" and ``wasserstein\_distance" of the \proglang{C++} library \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}.

>>>>

\author{\parbox{0.7\linewidth}{Brittany T. Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria, David L. Millman, \and Vincent Rouvreau} \\
  \small{In collaboration with the CMU TopStat Group}}
\Plainauthor{Brittany Terese Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria, David L. Millman, Vincent Rouvreau}

\title{Introduction to the \proglang{R} package \pkg{TDA}}
\Plaintitle{Introduction to the R package TDA}
\Abstract{
We present a short tutorial and introduction to using the \proglang{R} package 
\pkg{TDA}, which provides some tools for Topological Data Analysis.
In particular, it includes implementations of functions that, given some data, provide 
topological information about the underlying space, such as the distance function, the distance to a measure,
the kNN density estimator, the kernel density estimator, and the kernel distance.
The salient topological features of the sublevel sets (or superlevel sets) of these functions
can be quantified with persistent homology. We provide an \proglang{R} interface
for the efficient algorithms of the \proglang{C++} libraries \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, and \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}, 
including a function for the persistent homology of the Rips filtration, 
and one for the persistent homology of sublevel
sets (or superlevel sets) of arbitrary functions evaluated over a grid of points.
The significance of the features in the resulting persistence diagrams can be 
analyzed with functions that implement the methods discussed in \cite{fasy2014statistical},
\cite{chazal2014stochastic} and \cite{chazal2014robust}.
The \proglang{R} package \pkg{TDA} also includes the implementation of an algorithm for density 
clustering, which allows us to identify the spatial organization of the probability mass associated to 
a density function and visualize it by means of a dendrogram, the cluster tree. 
}
\Keywords{Topological Data Analysis, Persistent Homology, Density Clustering}

@@

  Jisu Kim\\
  DataShape Group\\
  INRIA Saclay\\
  Website: \url{ http://www.stat.cmu.edu/~jisuk/ } \\
  Email: \email{jisu.kim@inria.fr}\\

@@

  David L. Millman\\
  Computer Science Department\\
  Montana State University\\
  Website: \url{ http://millman.us } \\
  Email: \email{david.millman@montana.edu}\\

@@

%There is a growing list of software available to compute persistent homology.
%\textsf{\textbf{javaPlex}}\footnote{https://code.google.com/p/javaplex/} is a
%Java software package developed by the Computational Topology workgroup at
%Stanford University;
%\textsf{\textbf{phom}}\footnote{
%http://cran.r-project.org/web/packages/phom/index.html} is an R package written
%by Andrew Tausz;
%\textsf{\textbf{PHAT}}\footnote{https://bitbucket.org/phat-code/phat/} is a C++
%library developed at the Institute of Science and Technology Austria;
%\textsf{\textbf{Dionysus}}\footnote{https://www.mrzv.org/software/dionysus/} is
%a
%C++ library written and maintained by Dmitriy Morozov;
%finally, \textsf{\textbf{GUDHI}}\footnote{https://project.inria.fr/gudhi/} is
%new born project hosted by INRIA and whose goal is the development and
%implementation of new algorithms for geometric understanding in high
%dimensions.
%Preliminary results show that \textsf{\textbf{GUDHI}} outperforms its major
%competitors \textsf{\textbf{PHAT}} and \textsf{\textbf{Dionysus}}; see
%\cite{boissonnat2013compressed}.

This paper is devoted to the presentation of the \proglang{R} package \pkg{TDA},
which provides a user-friendly interface for the efficient algorithms of the 
\proglang{C++} libraries \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}} \citep{gudhi_stpcoh}, \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}} \citep{dionysus_C}, and \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}} \citep{phat_C}.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Persistent Homology Over a Grid}
%%%%%%%%%%%%%%%%%%%%%%%
In this section, we describe how to use the \code{gridDiag} function to compute the persistent homology of sublevel (and superlevel) sets of the functions described in Section \ref{sec:distances}. 
The function \code{gridDiag} evaluates a given real valued function over a triangulated grid, constructs a filtration of simplices using the values of the function, and computes the persistent homology of the filtration. From version 1.2, \code{gridDiag} works in arbitrary dimension.
The core of the function is written in \proglang{C++} and the user can choose to compute persistence diagrams using either  the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}.

@@

The \code{ripsDiag} function computes the persistence diagram of the Rips filtration built on top of a point cloud.
The user can choose to compute the Rips filtration using either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}} or \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}. Then for computing the persistence diagram from the Rips filtration, the user can use either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}.
\\The following code generates 60 points from two circles:

@@

The \code{alphaComplexDiag} function computes the Alpha complex filtration built on top of a point cloud, using the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}. Then for computing the persistence diagram from the Alpha complex filtration, the user can use either the C++ library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}.

@@

The \code{alphaShapeDiag} function computes the persistence diagram of the Alpha shape filtration built on top of a point cloud in 3 dimension, using the C++ library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}. Then for computing the persistence diagram from the Alpha shape filtration, the user can use either the C++ library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}. Currently the point data cloud should lie in 3 dimension.

@@

\noindent Once the filtration is computed, the function \code{filtrationDiag} computes the persistence diagram from the filtration. The user can choose to compute the persistence diagram using either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}} or \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}.

@@

Standard metrics for measuring the distance between two persistence diagrams are the bottleneck distance and the $p$th Wasserstein distance \citep{edelsbrunner2010computational}. 
The \pkg{TDA} package includes the functions \code{bottleneck} and \code{wasserstein}, which are R wrappers of the functions ``bottleneck\_distance" and ``wasserstein\_distance" of the \proglang{C++} library \href{ https://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}.
